---
title: "Music and Chronic Pain - Rmarkdown Final"
author: "Sydney Drake"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# QUESTION

Does music decrease the experience of chronic pain associated with fibromyalgia? What type of music has the greatest effect on chronic pain?

# HYPOTHESIS

Participants with fibromyalgia experiencing chronic pain will report less pain while listening to music. Classical music will have the most effective performance.

# STATISTICAL TEST

We will test the significance of our data using Repeated-Measures ANOVA, which will compare pain level across each of the individual's four conditions: initial pain, pain after pink noise, as well as pain before and after music of choice.

For visualization, we will compare the bar plot of individual's pain scores across these three conditions.

```{include=FALSE}
# Load libraries
library(tidyverse)
library(ez)

# Import data
data <- read.csv("dataset.csv")

# Check the structure of your dataset
str(data)
head(data)

# Reshape data from wide to long format
# (assuming each row = one participant, and columns = conditions)
data_long <- data %>%
  pivot_longer(cols = c(pic1, pic2, pi1, pim2),
               names_to = "Condition",
               values_to = "Pain")

# Add a participant ID if not already present
if(!"Participant" %in% names(data_long)){
  data_long$Participant <- rep(1:(nrow(data_long)/4), each = 4)
}

# Run repeated-measures ANOVA
anova_result <- ezANOVA(
  data = data_long,
  dv = Pain,
  wid = Participant,
  within = Condition,
  detailed = TRUE
)

# View results
print(anova_result)

# Optional: pairwise comparisons (post-hoc)
pairwise.t.test(data_long$Pain, data_long$Condition, paired = TRUE, p.adjust.method = "bonferroni")

# Optional: bar plot with error bars
ggplot(data_long, aes(x = Condition, y = Pain, fill = Condition)) +
  stat_summary(fun = mean, geom = "bar", color = "black", width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  theme_minimal() +
  labs(title = "Pain Ratings Across Conditions", y = "Pain (1–10)", x = "Condition")

```

# VISUALIZATION - BAR PLOT GRAPH

In the x-axis of the bar plot are 4 conditions. pic1 is the first control, which measures pain level before playing pink music, and pic2 measures pain level after the pink music. pim1 and pim2 show pain levels before and after playing music of choice respectively.

```{r boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make box plot work
data <- read.csv("dataset.csv")
data_long <- data %>%
  pivot_longer(cols = c(5, 6,7,8), names_to = "treatment", values_to = "value")

#plot
    ggplot(data_long, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  
    theme_classic() +
    theme(legend.position="none") +
    scale_fill_brewer(palette="Spectral") +
    theme(axis.text.x = element_text(size=15)) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    ylab("Pain Level") +
    xlab("")
```
    

# REFERENCES     

1. ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.
2. R-graph gallery. Used as a template for bar plot graph coding. Accessed 10/30/2025.


