---
title: "Warm-up mini-Report: Mosquito Blood Hosts in \n Salt Lake City, Utah"
author: "Sydney Drake"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

Our study focuses on the relationship between several species of Utah birds and the West Nile Virus (WNV), an endemic disease in the United States. Avian species act as hosts, carrying the disease in their blood, and potentially passing it onto humans through mosquitoes. In this study, mosquitoes were trapped and collected at several locations, and DNA was obtained through PCR and sequencing. This data allowed the identification of the species the mosquitoes were feeding on, allowing for an analysis on what species were being fed on the most. Additionally, statistic tests were run, testing the relationship between house finches and the presence and prevalence of the WNV to determine if there is a correlation between the house finch and the WNV. A positive correlation was found with a high significance, leading to a conclusion that there is a positive correlation between the two.

# BACKGROUND

The West Nile Virus (WNV) has been endemic in mainland United States since the late 1990s and has since achieved endemic status. Spread of this disease has been linked to mosquitoes and often carried by birds. Specific species of birds have been shown to have high viremia of the WNV when infected, making them a common host and link in human infection with this disease. Notably, the house finch is believed to be particularly susceptible to this disease, and has a high potential of passing it onto mosquitoes.

This study is focused on the link between House Finches and human WNV cases. Mosquito blood meal was taken from several locations, and data was compiled on those carrying the WNV and from what animal they took blood from. With the blood samples, PCR was performed and the resulting DNA was sequenced.
Our hypothesis addresses and quantifies the extent of WNV cases in House Finches specifically, and their likelihood of passing the disease onto humans.


```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
duration <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# Choose some colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(duration$mean, horiz=TRUE, names.arg=duration$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars
arrows(duration$lo, bp, duration$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Questions  
The question we will study and address is "What bird species is acting as WNV amplifying host in Salt Lake City?" as well as "To what degree are house finches and prevalence of the WNV correlated?"

## Hypothesis 
Our hypothesis is that house finches are acting as important amplifying hosts of WNV in Salt Lake City. We hypothesize that this specific species is particularly susceptible to spreading this disease as it produces a high viremia in it's blood.
  
## Prediction
We predict that, in the areas where mosquitoes commonly feed on house finches, trapped mosquitoes will have a higher rate of confirmed WNV in their systems.

# METHODS
First, mosquitoes were collected from several locations throughout the state of Utah using traps such as gravid traps and CO2 traps. The blood meal was run through PCR and sequenced. The results were compared to the BLAST database, identifying the species of blood meal within the mosquitoes. The resulting data were run through several statistical tests, of which are shown in this project, to determine significance. Additionally, graphs were compiled to visualize this data, as shown above in this project.

## First Analysis

The plot on the left shows the number of bloodmeals that came from each host species and locations were the WNV was not detected.
The graph on the right also shows the number of bloodmeals that came from each host species, though at locations where the WNV was detected.
This allows us to directly compare feeding patterns between infected and uninfected locations.

```{r first-analysis, warning=FALSE}
## import counts_matrix: data.frame with column 'loc_positives' (0/1) and host columns 'host_*'
counts_matrix <- read.csv("./bloodmeal_plusWNV_for_BIOL3070.csv")

## 1) Identify host columns
host_cols <- grep("^host_", names(counts_matrix), value = TRUE)

if (length(host_cols) == 0) {
  stop("No columns matching '^host_' were found in counts_matrix.")
}

## 2) Ensure loc_positives is present and has both levels 0 and 1 where possible
counts_matrix$loc_positives <- factor(counts_matrix$loc_positives, levels = c(0, 1))

## 3) Aggregate host counts by loc_positives
agg <- stats::aggregate(
  counts_matrix[, host_cols, drop = FALSE],
  by = list(loc_positives = counts_matrix$loc_positives),
  FUN = function(x) sum(as.numeric(x), na.rm = TRUE)
)

## make sure both rows exist; if one is missing, add a zero row
need_levels <- setdiff(levels(counts_matrix$loc_positives), as.character(agg$loc_positives))
if (length(need_levels)) {
  zero_row <- as.list(rep(0, length(host_cols)))
  names(zero_row) <- host_cols
  for (lv in need_levels) {
    agg <- rbind(agg, c(lv, zero_row))
  }
  ## restore proper type
  agg$loc_positives <- factor(agg$loc_positives, levels = c("0","1"))
  ## coerce numeric host cols (they may have become character after rbind)
  for (hc in host_cols) agg[[hc]] <- as.numeric(agg[[hc]])
  agg <- agg[order(agg$loc_positives), , drop = FALSE]
}

## 4) Decide species order (overall abundance, descending)
overall <- colSums(agg[, host_cols, drop = FALSE], na.rm = TRUE)
host_order <- names(sort(overall, decreasing = TRUE))
species_labels <- rev(sub("^host_", "", host_order))  # nicer labels

## 5) Build count vectors for each panel in the SAME order
counts0 <- rev(as.numeric(agg[agg$loc_positives == 0, host_order, drop = TRUE]))
counts1 <- rev(as.numeric(agg[agg$loc_positives == 1, host_order, drop = TRUE]))

## 6) Colors: reuse your existing 'cols' if it exists and is long enough; otherwise generate
if (exists("cols") && length(cols) >= length(host_order)) {
  species_colors <- setNames(cols[seq_along(host_order)], species_labels)
} else {
  species_colors <- setNames(rainbow(length(host_order) + 10)[seq_along(host_order)], species_labels)
}

## 7) Shared x-limit for comparability
xmax <- max(c(counts0, counts1), na.rm = TRUE)
xmax <- if (is.finite(xmax)) xmax else 1
xlim_use <- c(0, xmax * 1.08)

## 8) Plot: two horizontal barplots with identical order and colors
op <- par(mfrow = c(1, 2),
          mar = c(4, 12, 3, 2),  # big left margin for species names
          xaxs = "i")           # a bit tighter axis padding

## Panel A: No WNV detected (loc_positives = 0)
barplot(height = counts0,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "(-) WNV Locations",
        xlim = xlim_use)

## Panel B: WNV detected (loc_positives = 1)
barplot(height = counts1,
        names.arg = species_labels, 
        cex.names = .5,
        cex.axis = .5,
        col = rev(unname(species_colors[species_labels])),
        horiz = TRUE,
        las = 1,
        xlab = "Bloodmeal counts",
        main = "(+) WNV Locations",
        xlim = xlim_use)

par(op)

## Keep the colors mapping for reuse elsewhere
host_species_colors <- species_colors
```
## Second analysis/plot e.g. generalized linear model

We statistically evaluated whether the presence or number of house finch blood meals was associated with the WNV activity across collection sites. Specifically, we tested whether these variables predicted (1) the occurrence of WNV-positive mosquito pools (binary response: presence or absence) and (2) the WNV positivity rate at each site (continuous response). A logistic regression model was used for the binary outcome, and a linear regression model was used for the continuous outcome. These analyses allowed us to assess the relationship suggested by the exploratory bar plots and assess whether house finch feeding patterns were significantly associated with WNV occurrence and intensity across sites. 

```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```
```

## Second analysis/plot
```{r}
# second-analysis-or-plot, glm with house finch alone against binary +/_
glm1 <- glm(loc_positives ~ host_House_finch,
            data = counts_matrix,
            family = binomial)
summary(glm1)

#glm with house-finch alone against positivity rate
glm2 <- glm(loc_rate ~ host_House_finch,
            data = counts_matrix)
summary(glm2)
```

# DISCUSSION

With the collected data and statistical analyses, a clear positive relationship is seen between the population density of house finches and the presence and prevalence of the WNV.

This leads us to believe that house finches are susceptible carriers to the WNV and have the potential to exponentially increase the presence of WNV. This relationship is likely due to the high viremia in the blood of infected house finches. Further study may be beneficial as to why this species is particularly suseptible, and if any mechanisms exist in decreasing the suseptibility of house finches to the WNV.

## Interpretation - fill in analysis

Under binary statistical test, a p-value was calculated; 0.0287, which shows a high degree of confidence in the correlation. This leads us to believe that the presence of house finches increase the likelihood of positive WNV presence in a specific area.

## Interpretation - fill in analysis/plot

Following the continuous model, we can conclude that there is a positive correlation between the presence of house finches and the prevalence of WNV. This correlation is determined to be highly significant with a p-value of 0.000045. In other words, with more house finches present in an area, the WNV will become increasingly more common in trapped mosquitoes.

# CONCLUSION

In conclusion, our original hypothesis was proven correct with highly significant p-values from a continuous and binary statstical test. Due to these results, we can conclude that the presence of house finch populations positively correlates with the WNV, in both presence and the extent and spread of the disease. This has the potential to assist in future control of the spread of the WNV. Mosquito control resources can potentially be focused on these specific areas, which would thereby protect human populations.

# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. ChatGPT. OpenAI, version 5. Used to correct syntax errors. Accessed 10/10/2025.